MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-14 docs/paper/out/paper.tex
remove section about sample data -- paper
add workflow description from @rtobes -- paper
See https://github.com/ohnosequences/metagenomica/issues/26#issuecomment-140018675

\subsection{Pipeline Description}\label{pipeline-description}

@@

@@

For each read:

­1. Select only one BLASTN alignment (HSP) per reference sequence (the
HSP with lowest e value) 2. Filter all the HSPs with bitscore below a
defined BLASTN bitscore threshold s\_0 3. Find the best bitscore value S
in the set of BLASTN HSPs corresponding to hits of that read 4. Filter
all the alignments with bitscore below p * S (where p is a fixed by the
user coefficient to define the bitscore required, e.g.~if p=0.9 and
S=700 the required bitscore threshold would be 630) 5. Select all the
taxonomic nodes to which map the reference sequences involved in the
selected HSPs: - If all the selected taxonomic nodes forms a line in the
taxonomy tree (are located in a not branched lineage to the tree root)
we should choose the most specific taxID as the final assignment for
that read - If not, we should search for the (sensu stricto) Lowest
Common Ancestor (LCA) of all the selected taxonomic nodes (See Figure X)

In this approach the value used for evaluating the similarity is the
bitscore that is a value that increases when similarity is higher and
depends a lot on the length of the HSP

@@

We have maintained the simpler method of Best BLAST Hit (BBH) taxonomic
assignment because, in some cases, it can provide information about the
sequences that can be more useful than the obtained using LCA algorithm.
Using LCA algorithm when some reference sequences with BLAST alignments
over the required thresholds map to a not sufficiently specific taxID,
the read can be assigned to an unspecific taxon near to the root. If the
BBH reference sequence maps to a more specific taxa this method, in that
case, gives us useful information.

\subsection{Using MG7 with some example
data-sets}\label{using-mg7-with-some-example-data-sets}

We selected the datasets described in {[}Kennedy-2014{]} (??)

>>>>

\subsection{Workflow Description}\label{workflow-description}

The MG7 analysis workflow is summarized in Figure X. The input files for
MG7 are the FASTQ files resulting from a paired-end NGS sequencing
experiment.

\subsubsection{Joining reads of each pair using
FLASH}\label{joining-reads-of-each-pair-using-flash}

In the first step the paired-end reads, designed with an insert size
that yields pairs of reads with an overlapping region between them, are
assembled using FLASH \citep{magovc2011flash}. FLASH is designed to
merge pairs of reads when the original DNA fragments are shorter than
twice the length of reads. Thus, the sequence obtained after joining the
2 reads of each pair is larger and has better quality since the sequence
at the ends of the reads is refined merging both ends in the assembly.
To have a larger and improved sequence is crucial to do more precise the
inference of the bacterial origin based on similarity with reference
sequences.

\subsubsection{Parallelized BLASTN of each read against the
16S-DB7}\label{parallelized-blastn-of-each-read-against-the-16s-db7}

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database. Assignment based on direct
similarity of each read one by one compared against a sufficiently wide
database is a very exhaustive method for assignment
\citep{segata2013computational} . Some methods of assignment compare the
sequences only against the available complete bacterial genomes or avoid
computational cost clustering or binning the sequences first, and then
doing the assignments only for the representative sequence of each
cluster. MG7 carries out an exhaustive comparison of all the reads under
analysis and it does not applies any binning strategy. Every read is
specifically compared with all the sequences of the 16S database. We
select the best BLAST hits (10 hits by default) obtained for each read
to do the taxonomic assignment.

@@

All the reads are assigned under two different algorithms of assignment:
i. Lowest Common Ancestor based taxonomic assignment (LCA) and ii. Best
BLAST Hit based taxonomic assignment (BBH). Figure X displays
schematically the LCA algorithm applied sensu stricto (left panel) and
the called `in line' exception (right panel) designed in order to gain
specificity in the assignments in the cases in which the topology of the
taxonomical nodes corresponding to the BLAST hits support sufficiently
the assignment to the most specific taxon.

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a threshold of similarity (by default \(evalue \leq e^{-15}\))
filtering those hits that are not sufficiently good comparing them with
the best one. We select the best HSP (High Similarity Pair) per
reference sequence and then choose the best HSP (that with lowest e
value) between all the selected ones. The bitscore of this best HSP
(called S) is used as reference to filter the rest of HSPs. All the HSPs
with bitscore below p x S are filtered. p is a coefficient fixed by the
user to define the bitscore required, e.g.~if p=0.9 and S=700 the
required bitscore threshold would be 630. Once we have the definitive
HSPs selected, we obtain their corresponding taxonomic nodes using the
taxonomic assignments that NCBI provides for all the nt database
sequences. Now we have to analyze the topological distribution of these
nodes in the taxonomy tree: i. If all the nodes forms a line in the
taxonomy tree (are located in a not branched lineage to the tree root)
we should choose the most specific taxID as the final assignment for
that read. We call to this kind of assignment the `in line' exception
(see Figure X right panel). ii. If not, we should search for the
\emph{sensu stricto} Lowest Common Ancestor (LCA) of all the selected
taxonomic nodes (See Figure X left panel). In this approach we decided
to use the bitscore for evaluating the similarity because it is a value
that increases when similarity is higher and depends a lot on the length
of the HSP. Some reads could not find sequences with enough similarity
in the database and then they would be classified as reads with no hits.
Advanced metagenomics analysis approaches \citep{huson2013microbial}
have adopted LCA assignment algorithms because it provides fine and
trusted taxonomical assignment.

@@

We decided to maintain the simpler method of Best BLAST Hit (BBH) for
taxonomic assignment because, in some cases, it can provide information
about the sequences that adds information to that obtained using LCA
algorithm. Using LCA algorithm, when some reference sequences with BLAST
alignments over the required thresholds map to a not sufficiently
specific taxID, the read can be assigned to an unspecific taxon near to
the root of the taxonomy tree. If the BBH reference sequence maps to
more specific taxa, this method, in that case, gives us useful
information.

\subsubsection{Output for LCA and BBH
assignments}\label{output-for-lca-and-bbh-assignments}

MG7 provides independent results for the 2 different approaches, LCA and
BBH. The output files include, for each taxonomy node (with some read
assigned), abundance values for direct assignment and cumulative
assignment. The abundances are provided in counts (absolute values) and
in percentage normalized to the number of reads of each sample. Direct
assignments are calculated counting reads specifically assigned to a
taxonomic node, not including the reads assigned to the descendant nodes
in the taxonomy tree. Cumulative assignments are calculated including
the direct assignments and also the assignments of the descendant nodes.
For each sample MG7 provides 8 kinds of abundance values: LCA direct
counts, LCA cumu. counts, LCA direct \%, LCA cumu. \%, BBH direct
counts, BBH cumu. counts, BBH direct \%, BBH cumu. \%.
